---
title: "InPreD_README"
author: "Jaime Abraham Castro-Mondragon"
date: 'Last update: `r Sys.Date()`'
output: 
  html_document:
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: false
    self_contained: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r functions, include=FALSE}

insert.fig.toggle <- function(
  fig.id,        # Unique id for the div/button (e.g. "figVMWare")
  button.label,  # Label for the button (e.g. "Toggle Figure 1")
  img.src,       # Path to image (e.g. "fig/01_VMWare_login.png")
  fig.width = "45%",  # Width style for image
  fig.cap = "Caption text"  # Figure caption text
) {
  safe_fig_id <- gsub("'", "\\'", fig.id)
  safe_button_label <- gsub("'", "\\'", button.label)
  safe_img_src <- gsub("'", "\\'", img.src)
  safe_fig_cap <- gsub("'", "\\'", fig.cap)
  
  cat(sprintf('
<div style="margin-bottom: 10px; text-align: right;">
  <button class="btn-toggle" onclick="toggleFigure(\'%s\')">%s</button>
</div>
<div id="%s" style="display:block; text-align: center; margin-bottom: 20px;">
  <img src="%s" style="width:%s;" class="img-fluid"/>
  <p><strong>%s.</strong> %s</p>
</div>
', safe_fig_id, safe_button_label, safe_fig_id, safe_img_src, fig.width, safe_button_label, safe_fig_cap))
}

```


## Intro

This document explains how to run the InPreD pipeline to detect CNAs, driver mutations, gene fusions and other mutations from a Illumina TS0500 panel. Every week we sequence a batch of max 8 samples (hereafter referred as *sequencing run*). 

In this document we will use the following sequence run ID, we can easily identify by the first 6 digits corresponding to the data when the sequences were generated.

```{bash, echo=TRUE, eval = FALSE}
250515_AAAAAAAA_BBBB_CCCCCCCCCC
```

<!-- # Steps -->

<br>

## Connect to the Virtual Machines

<br>

1. In your browser, open the [VMWare Horizon](https://view.tsd.usit.no/portal/webclient/#/home) website. Log in with your *username* and *2FA* first, then VMWare will ask your *password* (Figure 1).

```{r figVMWare, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "figVMWare",
  button.label = "Figure 1",
  img.src = "fig/01_VMWare_login.png",
  fig.width = "45%",
  fig.cap = "Log in in the VMWare Horizon portal"
)

```

2. Click on the **p1150-rhel8-01-pool** icon, this will open the Linux VM (Figure 2). The other icon (*p1150-win02-pool*) will open the Windows VM.

```{r figVMs, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "figVMs",
  button.label = "Figure 2",
  img.src = "fig/02_VMs.png",
  fig.width = "45%",
  fig.cap = "Available VMs. Left: Linux, Right: Windows"
)

```

<br>

## Folder structure

3. We will be working in the following directory: `/ess/p1150/data/durable`. Next we have a list of folders that will be constantly used during the analysis, from now on, **the file paths are relative to the working directory**.


```{bash, echo=TRUE, eval = FALSE}
cd /ess/p1150/data/durable
```

 <br>
 
  - **`file-import/p1150-member-group/`** : this is the folder where raw files are available. The files are usually uploaded every friday.
  
<br>

  - **`file-export/`** : this is the folder where files can be exported outside the [TSD](https://selfservice.tsd.usit.no/) framework.

<br>

  - **`TSO500_SampleSheets/`** : in this folder we should place the Sample sheet files (see Step X) generated in every sequence run. Every sequence run should have three associated files. 
    - `250515_info.csv` : Original Samplesheet (for this sequencing run ID). 
    - `250515_info_TSOPPI.csv` : Customized Samplesheet (required by TSOPPI, may contain samples from many run IDs).
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC.csv` : Curated Samplesheet.

<br>

  - **`/TSO500_pronto_metadata/`** : in this folder we should place the metadata tables (original and in a compressed version; see Step X). Every sequence run should have two associated files. 
    - `PRONTO_metadata_250515.txt` : Original metadata. 
    - `PRONTO_metadata_250515.7z` : Compressed table.

<br>

  - **`TSO500_sequences/`** : in this folder we should place the sequence files (see Step X) generated in every sequence run. Every sequence run should have two associated files. 
    - `250515/` : Folder with sequences. 
    - `250515.md5` : File with checksums.

<br>

  - **`/TSO500_scripts_and_logs/`** : in this folder we will place the log and error messages of each analysis launched in the pipeline. One folder per sequence run ID.

<br>

  - **`TSO500_analyses/`** : here we will find the results after running *LocalApp*. Every sequence run should have one associated folder. 
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_LocalApp_analysis/` : this is a link. 
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_LocalApp_analysis_fromRAW/` : *LocalApp* results : 
        - `results/${SAMPLE_ID}/` : Sample ID results are in individual folders.
        - `results/MetricsOutput.tsv` : Table with sequencing metrics

<br>

  - **`TSO500_postprocessing/`** : here we will find the results after running *LocalApp* and *TSOPPI*, two folders per sequence run ID. 
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_LocalApp_postprocessing_results/` : *LocalApp*. 
      - `${SAMPLE_ID}` : One forlder per sample.
      - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_metrics_plots` : Sequencing QC-related plots
      - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_variant_summary.csv` : Table with variants
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_TSOPPI_v0.3.2_main_2022_06_07/`: *TSOPPI*

<br>

```{r toggle, echo=FALSE, results='asis'}

cat('
<style>
.btn-toggle {
  background-color: #0275d8;
  border: none;
  color: white;
  padding: 8px 16px;
  font-size: 14px;
  border-radius: 12px;       /* Rounded corners */
  cursor: pointer;
  transition: background-color 0.3s ease;
  box-shadow: 0 2px 4px rgba(0,0,0,0.15);
}
.btn-toggle:hover {
  background-color: #025aa5;
}
</style>

<script>
function toggleFigure(id) {
  var fig = document.getElementById(id);
  if (fig.style.display === "none" || fig.style.display === "") {
    fig.style.display = "block";
  } else {
    fig.style.display = "none";
  }
}
</script>
')
```


<!-- sequences -->
<!-- analysis -->
post-processing
icnva
export_for_interpretation
software
