---
title: "InPreD_README"
author: "Jaime Abraham Castro-Mondragon"
date: 'Last update: `r Sys.Date()`'
output:
  html_document:
    number_sections: true
    theme: cosmo
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
    self_contained: true
    code_folding: show
---

```{r setup, include=FALSE}
library(emo)
library(glue)
library(broom)
library(klippy)

knitr::opts_chunk$set(echo = TRUE)

date.checklist <- substr(gsub(x = Sys.Date(), pattern = "-", ""), start = 3, 8)

checklist.terms <- list()
run.id          <- NULL


# Force user to define a Run ID
run.id <- "250620_AAAAAAAA_BBBB_CCCCCCCCCC" # "250606_AAAAAAAA_BBBB_CCCCCCCCCC"
if (is.null(run.id)){
  stop("run.id cannot be NULL")
}

run.date <- unlist(strsplit(run.id, split = "_"))[1]

if (file.exists("InPreD_README.html")) {
  file.remove("InPreD_README.html")
}


checklist.file <- paste0(run.date, "_InPreD_pipeline_Checklist.html")
if (file.exists(checklist.file)) {
  file.remove(checklist.file)
}
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(lang     = c('r', 'python', 'bash'),
               position = c('top', 'right'),
               color    = '#253494')
```

```{r functions, include=FALSE}
# ----------------------- #
# Show/Hide figure botton #
# ----------------------- #

insert.fig.toggle <- function(
  fig.id,        # Unique id for the div/button (e.g. "figVMWare")
  button.label,  # Label for the button (e.g. "Toggle Figure 1")
  img.src,       # Path to image (e.g. "fig/01_VMWare_login.png")
  fig.width = "45%",  # Width style for image
  fig.cap = "Caption text"  # Figure caption text
) {
  safe_fig_id <- gsub("'", "\\'", fig.id)
  safe_button_label <- gsub("'", "\\'", button.label)
  safe_img_src <- gsub("'", "\\'", img.src)
  safe_fig_cap <- gsub("'", "\\'", fig.cap)
  
  cat(sprintf('
<div style="margin-bottom: 10px; text-align: right;">
  <button class="btn-toggle" onclick="toggleFigure(\'%s\')">%s</button>
</div>
<div id="%s" style="display:block; text-align: center; margin-bottom: 20px;">
  <img src="%s" style="width:%s;" class="img-fluid"/>
  <p><strong>%s.</strong> %s</p>
</div>
', safe_fig_id, safe_button_label, safe_fig_id, safe_img_src, fig.width, safe_button_label, safe_fig_cap))
}

# ------------------------------ #
# Checklist names and categories #
# ------------------------------ #
make_prepender <- function(prefix_name) {
  function(label, v) {
    c(label, v)
  }
}


add.cl.section    <- make_prepender("section")
add.cl.categories <- make_prepender("category")
```



```{r checklist, include=FALSE}

checklist.template.file <- "templates/InPreD_pipeline_Checklist.html"
checklist.ready.file    <- paste0(run.date, "_InPreD_pipeline_Checklist.html")

# ------------------------------- #
# Prepare CheckList section names #
# ------------------------------- #

# CheckList section names
cl.sections   <- vector()
cl.categories <- vector()

# Check List item
generate_checklist_html <- function(checklist, categories) {
  if (length(checklist) != length(categories)) {
    stop("Checklist and categories must have the same length.")
  }
  
  html_items <- mapply(function(item, i, category) {
    sprintf(
      '<div class="checklist-item %s">\n  <input type="checkbox" id="item%d">\n  <label for="item%d">%d. %s</label>\n</div>',
      category, i, i, i, item
    )
  }, checklist, seq_along(checklist), categories, USE.NAMES = FALSE)
  
  paste(html_items, collapse = "\n")
}


## Copy template and read copy
file.copy(from = checklist.template.file,
          to   = checklist.ready.file)

```

<br>

## Introduction

<br>

This document explains how to run the *InPreD* pipeline to detect CNAs, driver mutations, gene fusions and other mutations from a *Illumina TS0500* panel. Every week we sequence a batch of max 8 samples (hereafter referred as *sequencing run*), with its corresponding unique run ID.

<br>

```{r runID, echo=FALSE, eval = TRUE, comment=""}
cat(run.id)
```

<br>

## Connect to the Virtual Machines

<br>

  - In your browser, open the [View TSD Portal](https://view.tsd.usit.no/portal/webclient/#/home) website. Log in with your *username* and *2FA* first, then VMWare will ask your *password* (Figure 1).

```{r figVMWare, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "figVMWare",
  button.label = "Figure 1",
  img.src = "fig/01_VMWare_login.png",
  fig.width = "45%",
  fig.cap = "Log in in the VMWare Horizon portal"
)

```

  - Click on the **p1150-rhel8-01-pool** icon, this will open the Linux VM (Figure 2). The other icon (*p1150-win02-pool*) will open the Windows VM.

```{r figVMs, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "figVMs",
  button.label = "Figure 2",
  img.src = "fig/02_VMs.png",
  fig.width = "45%",
  fig.cap = "Available VMs. Left: Linux, Right: Windows"
)

```

<br>

## Folder structure

<br>

We will be working in the following directory: `/ess/p1150/data/durable`. We have a list of subfolders that will be constantly used during the analysis, from now on, **the file paths are relative to the working directory `durable`**.


```{bash, echo=TRUE, eval = FALSE}
cd /ess/p1150/data/durable
```

<br>
 
  - **`file-import/p1150-member-group/`** : all uploaded files through the [*TSD Data Portal*](data.tsd.usit.no) will be located here (e.g., sequencing raw files, sample sheets and metadata). The files are uploaded usually every friday afternoon. See section *4 Import Samplesheets to TSD*.

<br>

  - **`file-export/`** : this is the folder where files can be exported outside the [TSD](https://selfservice.tsd.usit.no/) framework. Any file in this folder is visible outside the *TSD* framework and can be downloaded, so we must be cautious about its content.

<br>

  - **`TSO500_SampleSheets/`** : in this folder we place the Sample sheet files (see Step X) generated in every sequence run. Every sequence run should have three associated files. 
    - ``r paste0(run.date, "_info.csv")`` : Original Samplesheet (for this run ID). Taken from `file-import/` folder, do not modify this file.
    - ``r paste0(run.date, "_info_TSOPPI.csv")`` : Copy of the original Samplesheet, it may be modified. This copy is required by *TSOPPI*, may contain samples from many run IDs or commented samples.
    - ``r paste0(run.id, ".csv")`` : Curated Samplesheet.

<br>

  - **`/TSO500_pronto_metadata/`** : in this folder we place the metadata tables (original and in an encrypted version; see Step X). Every sequence run should have two associated files. 
    - ``r paste0("PRONTO_metadata_", run.date, ".txt")`` : Original metadata table. 
    - ``r paste0("PRONTO_metadata_", run.date, ".7z")`` : Compressed and encrypted table.

<br>

  - **`TSO500_sequences/`** : in this folder we place the sequence files (see Step X) generated in every run. Every sequence run should have two associated files. 
    - ``r paste0(run.id, "/")`` : Folder with sequences. 
    - ``r paste0(run.id, ".md5")`` : File with checksums.

<br>

  - **`/TSO500_scripts_and_logs/`** : in this folder we will place the logs and error messages of each analysis launched in the pipeline. One folder per run ID.

<br>

  - **`TSO500_analyses/`** : here we will find the results after running *LocalApp*. Every sequence run should have one associated folder. 
    - ``r paste0(run.id, "_LocalApp_analysis/")`` : this is a link. 
    - ``r paste0(run.id, "_LocalApp_analysis_fromRAW/")`` : *LocalApp* results : 
        - `results/${SAMPLE_ID}/` : Sample ID results are in individual folders
        - `results/MetricsOutput.tsv` : Table with sequencing metrics

<br>

  - **`TSO500_postprocessing/`** : here we find the results after running *LocalApp* and *TSOPPI*, two folders per run ID. Both folders follow the same structure.
  
    - ``r paste0(run.id, "_TSO500_LocalApp_postprocessing_results/")`` : *LocalApp*. 
        - `${SAMPLE_ID}` : One folder per sample.
        - ``r paste0(run.id, "_metrics_plots")`` : Sequencing QC-related plots.
        - ``r paste0(run.id, "_variant_summary.csv")`` : Table with variants.

    - ``r paste0(run.id, "_TSO500_TSOPPI_v0.3.2_main_2022_06_07/")`` : *TSOPPI*.

<br>

  - **`TSO500_icnva/`** : here we find the [*PureCN*](https://scfbm.biomedcentral.com/articles/10.1186/s13029-016-0060-z) results. One folder per run ID, and one folder per sample ID.
  
    - ``r paste0(run.id, "/${SAMPLE_ID}_icnva/")``
        - `PureCN_results_default/` : *PureCN* results of each patient. Including the local optima solutions for ploidy/purity.
        - `PureCN_results_custom/` : *PureCN* results of each patient.
        - `${SAMPLE_ID}_NRC_plots.pdf` : Tumor vs Normal samples comparison plots.

<br>

  - **`TSO500_export_for_interpretation/`** : this is the folder where the results are stored before the interpretation of variants, mutations and fusions.
  
    - `RECENT_RUN/` : here we will find the 8 samples corresponding to the current sequencing run. The content is similar to `TSO500_icnva/`.
    - `IPA001-IPA100` : there are many folders with a similar pattern in their name. After interpretation the `RECENT_RUN/` content must be moved to their corresponding folder.

<br>

  - **`software/`** : here we find the bioinformatic analysis pipeline scripts and the source code for all the software required. 
    - **`pipeline/v003_upgrading_pronto_to_1.2.0`**

<br>

## Import Samplesheets to TSD

<br>

  - Before launching the pipeline, we must generate the SampleSheet and *PRONTO* metadata from the [View UiO TSD portal](https://view.uio.no/portal/webclient/#/home). Note that this is a *Windows* VM.

  - Log in into the [View UiO TSD portal](https://view.uio.no/portal/webclient/#/home). You must write your *2FA* one time code and your password.
  
  - Click on *Windows icon* -> *This PC* -> *Hypathia.uio.no* (Figure 3).
  
  - **SampleSheet** file: 
    
    - `hypathia.uio.no/Ih-med-klinmed-ahus-inpred/TS0500/Samples_YYYY_wWW.xlsx` (where *YYYY* = Year and *wWW* = week. Example: *w22* for week 22; see Figure 4).

```{r ViewUiO, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "Hypathia",
  button.label = "Figure 3",
  img.src = "fig/03_Hypatia.png",
  fig.width = "75%",
  fig.cap = "Access the Hypathia server thorugh UiO View Portal"
)

```

<br>

```{r ViewUiO_SampleSheet, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "ViewUiO_SampleSheet ",
  button.label = "Figure 4",
  img.src = "fig/04_SampleSheet.png",
  fig.width = "75%",
  fig.cap = "Original SampleSheet in View UiO portal"
)

```

<br>

### Generate *SampleSheet* and *PRONTO* metadata

<br>

  - Open the *SampleSheet file* with *Microsoft Excel*. You will see two sheets, namely **Sheet1** and **PRONTO**.

<br>

  - **SampleSheet**:
  
    - In *Sheet1*, select columns **InPred ID (IPA)** (E), **Tumor content** (R), and  **Barcode** (S).
    - Create a new file with these columns and  **remove the header**. The file must be named `SampleSheet_YYMMDD_info.csv`.
    
    - Save the new file in the metadata folder `\\platon.uio.no\med-klinmed-u2\jamondra\pc\Dokumenter\Metadata`.

<br>

  - **PRONTO metadata**:
  
    - In *PRONTO* sheet. Save the content as it is in a new **tab separated file (.txt)**. The file must be named `PRONTO_metadata_YYMMDD.txt`.
    
    - Save the new file in the metadata folder `\\platon.uio.no\med-klinmed-u2\jamondra\pc\Dokumenter\Metadata`.
    
    - **Encrypt this file in *7z* format.** *Right click* on the file and select the *7-zip* option, select the *Leff til arkiv* option (Figure 5). This will open a window, write and repeat your encryption password in the *Angi passord* and *Bekreft passord* text boxes (Figure 6). Do not modify the other parameters.

```{r Encrypt_PRONTO_pt1, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "Encrypt_PRONTO_pt1",
  button.label = "Figure 5",
  img.src = "fig/05_Encrypt_pt1.png",
  fig.width = "85%",
  fig.cap = "Right click on the PRONTO metadata file to display encryption options"
)

```

```{r Encrypt_PRONTO_pt2, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "Encrypt_PRONTO_pt2",
  button.label = "Figure 6",
  img.src = "fig/06_Encrypt_pt2.png",
  fig.width = "65%",
  fig.cap = "Write and repeat your encryption password"
)

```

<br>

  - **NOTE:** both *SampleSheet* and *PRONTO* metadata file **must** have the exact same date *YYMMDD*.
  

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Create SampleSheet (View UiO)", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Create PRONTO metadata (View UiO)", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Encrypt PRONTO metadata", cl.sections)
cl.categories <- add.cl.categories("important", cl.categories)
```

<br>

### Import *SampleSheet* and *PRONTO* metadata (Uio View -> TSD)

<br>

  - **NOTE**: the TSD import must be done through the *UiO View portal* only, never through a personal computer.
  
  - Open *Chrome* or any internet browser.
  
  - Go to the [*TSD / Data Portal*](data.tsd.usit.no) and click on *Import Files* button (Figure 7).
  
  - Log in with your *username*, *2F2 one time code*. and *password*.
  
  - Upload the `SampleSheet_YYMMDD.csv` and `PRONTO_metadata_YYMMDD.7z` files. Verify both files are listed and checked, and click on the *import button* (Figure 8).
  
  - **NOTE**: only import the encrypted *PRONTO metadata* file, never the *.txt (non-encrypted)* version.

<br>

```{r TSD_import, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "TSD_import",
  button.label = "Figure 7",
  img.src = "fig/07_TSD_DP.png",
  fig.width = "65%",
  fig.cap = "Import Files in TSD Data Portal"
)

```

<br>

```{r TSD_import_button, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "TSD_import_button",
  button.label = "Figure 8",
  img.src = "fig/08_TSD_import_button.png",
  fig.width = "85%",
  fig.cap = "Verify SampleSheet and encrypted PRONTO metadata are listed and checked. Then click on the import button"
)

```

<br>

  - Open the [View TSD Portal](https://view.tsd.usit.no/portal/webclient/#/home) with the *Windows VM*.
  
  - Verify that `SampleSheet_YYMMDD.csv` and `PRONTO_metadata_YYMMDD.7z` files are in the *file-import* folder (Figure 9): **`file-import/p1150-member-group/`**.
  
  - Right click on the `PRONTO_metadata_YYMMDD.7z` file, select the *7-zip* option and *Pakk ut ...* (Figure 10).
  
  - Write the encryption password and click on *OK* (Figure 11).
  
<br>

```{r TSD_import_files_windows, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "TSD_import_files_windows",
  button.label = "Figure 9",
  img.src = "fig/09_TSD_import-files.png",
  fig.width = "85%",
  fig.cap = "In the import-files folder you will see the SampleSheet and the encrypted PRONTO metadata"
)

```

<br>

```{r Decrytp_PRONTO_pt1, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "Decrytp_PRONTO_pt1",
  button.label = "Figure 10",
  img.src = "fig/10_Decrypt_pt1.png",
  fig.width = "85%",
  fig.cap = "Right click on the .7z file to show the decryption options."
)
```

<br>

```{r Decrytp_PRONTO_pt2, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "Decrytp_PRONTO_pt2",
  button.label = "Figure 11",
  img.src = "fig/11_Decrypt_pt2.png",
  fig.width = "85%",
  fig.cap = "Write the decryption password"
)
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Import files in TSD Data Portal", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Decrypt PRONTO metadata", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

## Prepare pipeline

<br>

Before launching the pipeline we must verify if there is enough space in the disk and the size of the sequencing data.

<br>

### Verify disk space availability

  - **Disk space** : we need at least **200GB of free space**. If this is not the case, contact people involved in the project to release some space.

```{bash, echo=TRUE, eval = FALSE}
# Check disk space
df -h /ess/p1150
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Verify disk space availability", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Verify size of sequencing run

  - **Size of files from current run ID**: usually a sequencing run produces a folder with size varying from **~30-60 GB**. If this number is outside this range, please contact the people involved in sequencing.

```{bash, echo=TRUE, eval = FALSE}
# Check size of sequencing results
du -csh /ess/p1150/data/durable/file-import/p1150-member-group/tsd
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Verify size of sequencing run", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Connect to hpc (*Colossus*) server.

  - **NOTE**: we must be connected to the *Colossus* server to launch the pipeline. Data manipulation can be done outside.

```{bash, echo=TRUE, eval = FALSE}
# Connect to hpc
ssh p1150-hpc-01
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Connect to hpc (Colossus) server", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Pipeline run directory.

  - **NOTE**: the pipeline must be launched from this folder.

```{bash, echo=TRUE, eval = FALSE}
cd  /ess/p1150/data/durable/software/pipeline/v003_upgrading_pronto_to_1.2.0
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Pipeline run directory", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Set up global variables.

<br>

The following variables must be set up before launching the pipeline:

  - Sequencing run ID = `RUN_ID`
  - Input data type = `INPUT_DATA_TYPE`
  - Sequencing run date = `RUN_DATE`
  - Durable folder = `DURABLE_DIR`
  
```{r env_var, echo=FALSE, eval = TRUE, comment=""}
cat(glue('RUN_ID="{run.id}"
RUN_DATE="${{RUN_ID%%_*}}"
INPUT_DATA_TYPE="RAW"
DURABLE_DIR="/tsd/p1150/data/durable"'), "\n")
```

```{bash, echo=TRUE, eval = FALSE}
# Check ENV variables content
echo $RUN_ID
echo $RUN_DATE
echo $INPUT_DATA_TYPE
echo $DURABLE_DIR
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Set up  ${RUN_ID}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Set up  ${RUN_DATA}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Set up  ${INPUT_DATA_TYPE}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Set up  ${DURABLE_DIR}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)
```

<br>

### Prepare sequences and metadata

<br>

The sequences, the *PRONTO* metadata, and the *SampleSheet* for the current run ID are located in the `file-import` folder.
They must be moved to the `TSO500_sequences`, `TSO500_pronto_metadata`, and `TSO500_SampleSheets` folders, respectively, before launching the analysis.

**NOTE**: An invisible character is inserted in the header of the `SampleSheet_$DATE.csv` metadata file. This character must be removed because the table cannot be properly read by other software and crash the analysis. Once this issue is solved, proceed to move the folders.

```{bash, echo=TRUE, eval = FALSE}
# Move sequences file
mv ${DURABLE_DIR}/file-import/p1150-member-group/tsd/${RUN_DATE}* ${DURABLE_DIR}/TSO500_sequences/
```

```{bash, echo=TRUE, eval = FALSE}
# Move SampleSheet metadata file
mv ${DURABLE_DIR}/file-import/p1150-member-group/SampleSheet_${RUN_DATE}_info.csv ${DURABLE_DIR}/TSO500_SampleSheets/SampleSheet_${RUN_DATE}_info.csv
```

<!-- We need to move two files: the .7z and .txt -->
```{bash, echo=TRUE, eval = FALSE}
# Move PRONTO file
mv ${DURABLE_DIR}/file-import/p1150-member-group/PRONTO_metadata_${RUN_DATE}* ${DURABLE_DIR}/TSO500_pronto_metadata/PRONTO_metadata_${RUN_DATE}*
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Remove invisible character in ShampleSheet file", cl.sections)
cl.categories <- add.cl.categories("important", cl.categories)

cl.sections   <- add.cl.section("Move sequences files", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Move ShampleSheet", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Move PRONTO metadata", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

### Verify flags and source `config.sh`

 - Verify the three following flags in the config file.
  
```{bash, echo=TRUE, eval = FALSE}
# Default ';' but other separator (',', '-', '/') can be used
SEPARATOR=";"

DECRYPT_UNPACK_CHECKSUM="T"
REMOVE_DECRYPTED_COMPRESSED_DATA="T"
```

<br>

 - Then source the `config` file

```{bash, echo=TRUE, eval = FALSE}
source config.sh ${RUN_ID}
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("config.sh : SEPARATOR", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("config.sh : DECRYPT_UNPACK_CHECKSUM", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("config.sh : REMOVE_DECRYPTED_COMPRESSED_DATA", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Source config.sh", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

```

<br>

### Prepare folder structure

<br>

 - Prepare the folder structure for a new run, this will create all folders to store
results and logs.

```{bash, echo=TRUE, eval = FALSE}
bash 00_setup_folder_structure_for_new_run.sh ${RUN_ID}
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Prepare folder structure", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)
```

<br>

### Generate SampleSheet (00)

<br>

  - We will generate a sample sheet with metadata for the current run ID.

```{bash, echo=TRUE, eval = FALSE}
bash 00_generate_TSO500_sample_sheet.sh ${RUN_ID} ${INPUT_DATA_TYPE} \
  &> ${SCRIPTS_AND_LOGS_FOLDER}/00_generate_TSO500_sample_sheet.log
```

<br>

  - Copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 00_generate_TSO500_sample_sheet.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/00_config.sh
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Generate sample sheet", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Copy Samplesheet (00) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

## Launch TSO500 pipeline

<br>

  - Now we are ready to launch the *TSO500* analysis pipeline. It has to be launched
in many steps, and many of them require manual intervention.

<br>

### Launch *LocalApp* analysis (01)

<br>

  - This analysis performs the data QC for the sequencing results.

`r emo::ji("hourglass_flowing_sand")` Running time: ~13 hours

```{bash, echo=TRUE, eval = FALSE}
bash 01_raw_sequencing_data_management_and_TSO500_analysis.sh ${RUN_ID} ${INPUT_DATA_TYPE} \
  > ${SCRIPTS_AND_LOGS_FOLDER}/01_raw_sequencing_data_management_and_TSO500_analysis.stdout \
  2> ${SCRIPTS_AND_LOGS_FOLDER}/01_raw_sequencing_data_management_and_TSO500_analysis.stderr &
```

<br>

  - **Send an email to the local *InPreD* team** to to tell them that the analyses have
been launched. You can find a draft in the *Email Drafts* section below in this
document.

<br>

  - Once this analysis is done, copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 01_raw_sequencing_data_management_and_TSO500_analysis.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/01_config.sh
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Launch LocalApp", cl.sections)
cl.categories <- add.cl.categories("analysis", cl.categories)

cl.sections   <- add.cl.section("Send LocalApp email", cl.sections)
cl.categories <- add.cl.categories("email", cl.categories)

cl.sections   <- add.cl.section("Copy LocalApp (01) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

### Launch *TSOPPI* analysis

<br>

This analysis performs different analysis included in the  [TSOPPI](https://tsoppi.readthedocs.io/en/latest/index.html) tool, such as:

 - Quality control (QC) metrics for samples
 - Fusion and splice variant output for RNA samples;
 - Small variant output, copy number variant output (CNV) and microsatellite status (MS status) information for DNA samples.

<br>

#### Create SampleSheet info *TSOPPI*

<br>

  - Create the *TSOPPI* samplesheet.

```{bash, echo=TRUE, eval = FALSE}
cp ${SAMPLESHEETS_FOLDER}/${INFO_FILE_NAME} ${SAMPLESHEETS_FOLDER}/${INFO_FILE_NAME_TSOPPI}
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Copy SampleSheet info TSOPPI", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

#### Adapt ALL_SEQUENCING_RUN_IDs global variable

<br>

  - Update the file `$ALL_RUN_IDS` and add the new `$RUN_ID` in the second line. See the example below.

```{bash, echo=TRUE, eval = FALSE}
ALL_SEQUENCING_RUN_IDs=()
#  -->    HERE       <---        # Insert new SEQ_ID in this line #
ALL_SEQUENCING_RUN_IDs+=( "250522_AAAAAAAA_BBBB_CCCCCCCCCC" )
ALL_SEQUENCING_RUN_IDs+=( "250515_AAAAAAAA_BBBB_CCCCCCCCCC" )
```

<br>

  - **NOTE**: originally this step was done manually, now we added the code below to do it automatically.

```{bash, echo=TRUE, eval = FALSE}
ALL_RUN_IDs="${PIPELINE_FOLDER}/all_sequencing_runs.sh"

# Backup copy
cp $ALL_RUN_IDS ${ALL_RUN_IDS}.bk

# Check if the RUN_ID is already present
if grep -qF "$RUN_ID" "$ALL_RUN_IDS"; then
    echo "RUN_ID already present in $ALL_RUN_IDS"
else
    # Insert RUN_ID after the initialization line
    TEMP_FILE=$(mktemp)
    insert_line='ALL_SEQUENCING_RUN_IDs+=( "'"$RUN_ID"'" )'

    awk -v line="$insert_line" '
        $0 ~ /^ALL_SEQUENCING_RUN_IDs=\(\)/ {
            print
            print line
            next
        }
        { print }
    ' "$ALL_RUN_IDS" > "$TEMP_FILE" && mv "$TEMP_FILE" "$ALL_RUN_IDS"
fi
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Update ${ALL_RUN_IDs} variable", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)
```

<br>

#### Launch *TSOPPI* (02)

<br>

`r emo::ji("hourglass_flowing_sand")` Running time: ~30 minutes

```{bash, echo=TRUE, eval = FALSE}
bash 02_execute_tsoppi_and_archive_results.sh ${RUN_ID} \
  > ${SCRIPTS_AND_LOGS_FOLDER}/02_execute_tsoppi_and_archive_results.stdout \
  2> ${SCRIPTS_AND_LOGS_FOLDER}/02_execute_tsoppi_and_archive_results.stderr &
```

<br>

  - Once this analysis is done, copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 02_execute_tsoppi_and_archive_results.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/02_config.sh
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Launch TSOPPI", cl.sections)
cl.categories <- add.cl.categories("analysis", cl.categories)

cl.sections   <- add.cl.section("Copy TSOPPI (02) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

### Launch *ICNVA* analysis

<br>

[*ICNVA*](https://tsoppi.readthedocs.io/en/latest/ICNVA.html) runs two types of analysis on each input sample:

  - Allele-specific copy number analysis performed by *PureCN* (LoH assessment.)
  - Normalized read coverage (NRC) profile plotting for selected genes (identify/confirm intra-genic duplications/losses).

  - An in-house set of controls is utilized during both analyses.

<br>

#### Select DNA tumour samples

<br>

  - Adapt the `$SAMPLE_IDs` array to contain only DNA tumour samples, those ones containing **-D01-** in the second field in their sample ID.

  - Be aware to adapt the variable according to the example below. Keep in mind the spacing, quotations and no separators.

```{bash, echo=TRUE, eval = FALSE}
SAMPLE_IDs=( "SAMPLE1" "SAMPLE_2" ... "SAMPLE_3" )
```

<br>

  - **NOTE**: originally this step was done manually, now we added the code below to do it automatically.

```{bash, echo=TRUE, eval = FALSE}
# Get sample IDs from SampleSheet
DNA_TUMOUR_SAMPLE_IDS=$(cut -d',' -f1 ${SAMPLESHEETS_FOLDER}/${INFO_FILE_NAME} | awk -F'-' '$2 == "D01"')
#DNA_TUMOUR_SAMPLE_IDS=$(cut -d',' -f1 samples.txt | awk -F'-' 'tolower($2) == "d01"') # Case insensitive alternative

# Build array declaration string
ARRAY_STRING="SAMPLE_IDs=("
while IFS= read -r id; do
    ARRAY_STRING+=" \"$id\""
done <<< "$DNA_TUMOUR_SAMPLE_IDS"
ARRAY_STRING+=" )"

# Evaluate the string to create the array variable
eval "$ARRAY_STRING"
```

<br>

  - Verify `SAMPLE_IDs` size and content.

```{bash, echo=TRUE, eval = FALSE}
# Optional: check the array content
declare -p SAMPLE_IDs

# Print array size and content in screen
echo ${#SAMPLE_IDs[@]}
printf "%s\n" "${SAMPLE_IDs[@]}"
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Update ${SAMPLE_IDs} variable", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Verify ${SAMPLE_IDs} size and content", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)
```

<br>

#### Launch *ICNVA* (025)

`r emo::ji("hourglass_flowing_sand")` Running time: ~30 minutes

```{bash, echo=TRUE, eval = FALSE}
for SAMPLE_ID in ${SAMPLE_IDs[@]}
do
      echo "SAMPLE_ID=${SAMPLE_ID}"
      echo "RUN_ID=${RUN_ID}"
      
      bash 025_run_initial_icnva.sh \
          config.sh                 \
          ${RUN_ID}                 \
          ${SAMPLE_ID}              \
          > ${SCRIPTS_AND_LOGS_FOLDER}/025_run_initial_icnva_${SAMPLE_ID}_${RUN_ID}.stdout \
          2> ${SCRIPTS_AND_LOGS_FOLDER}/025_run_initial_icnva_${SAMPLE_ID}_${RUN_ID}.stderr &
done
```

<br>

  - Once this analysis is done, copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 025_run_initial_icnva.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/025_config.sh
```

<br>

  - **Send an email to the local *InPreD* team** to to tell them that *TSOPPI* and *ICNVA* analyses are ready and they can proceed with QC. You can find a draft in the *Email Drafts* section below in this
document.

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Launch ICNVA", cl.sections)
cl.categories <- add.cl.categories("analysis", cl.categories)

cl.sections   <- add.cl.section("Copy ICNVA (025) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Send TSOPPI + ICNVA email", cl.sections)
cl.categories <- add.cl.categories("email", cl.categories)

```

<br>

#### Process QC variant tables (023)

<br>

Convert QC-ed small variables tables, from *xlsx* back to *tsv* format.

`r emo::ji("hourglass_flowing_sand")` Running time: ~3 minutes

```{bash, echo=TRUE, eval = FALSE}
for SAMPLE_ID in ${SAMPLE_IDs[@]}
do
      echo "SAMPLE_ID=${SAMPLE_ID}"
      echo "RUN_ID=${RUN_ID}"
      
      bash 023c_convert_variant_table_forQC_xlsx_to_tsv.sh \
          config.sh                 \
          ${RUN_ID}                 \
          ${SAMPLE_ID}              \
          > ${SCRIPTS_AND_LOGS_FOLDER}/023c_convert_variant_table_forQC_xlsx_to_tsv_${SAMPLE_ID}_${RUN_ID}.stdout \
          2> ${SCRIPTS_AND_LOGS_FOLDER}/023c_convert_variant_table_forQC_xlsx_to_tsv_${SAMPLE_ID}_${RUN_ID}.stderr &
done
```

<br>

  - Once this analysis is done, copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 023c_convert_variant_table_forQC_xlsx_to_tsv.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/023_config.sh
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Convert tables: xlsx -> tsv", cl.sections)
cl.categories <- add.cl.categories("analysis", cl.categories)

cl.sections   <- add.cl.section("Copy xlsx -> tsv (023) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

### Generate reports

<br>

Once the list of mutations was filtered by the curators, we are ready to proceed
and generate the reports using *PRONTO*.

<br>

#### Update *PRONTO* metadata master table

<br>

  - Update the *PRONTO* master table metadata: `${DURABLE}/software/PRONTO/PRONTO-1.2.0/In/InPreD_PRONTO_metadata.txt`
  
    - Comment previous run by adding a sharp `#` one line before the first entry of the previous run.
    
    - Set all values of the 3rd column (*Create report*) to *N*.
    
    - Add the values of current run ID PRONTO metadata to the master table after the header. Double check to not add the header twice.
    
    - Replace norsk characters:
      
      - Æ/æ : AE/ae
      - Ø/ø : OE/oe
      - Å/å : Aa/aa
    
    - Set all values of the 2nd columns (*Sequencing_run_id*) to the current run ID: ``r paste0(run.id)``.

```{bash, echo=TRUE, eval = FALSE}
#cp -a 023c_convert_variant_table_forQC_xlsx_to_tsv.sh ${SCRIPTS_AND_LOGS_FOLDER}
#cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/023_config.sh
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Comment previous run in PRONTO master table", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Set 'N' in 3rd column in PRONTO master table", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Add current ID PRONTO metadata to PRONTO master table", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Replace Æ/æ in PRONTO master table", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Replace Ø/ø in PRONTO master table", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Replace Å/å in PRONTO master table", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

#### Generate reports with PRONTO

<br>



  - Update PRONTO metadata master table: add the metadata of the current run.
  - Update column with Ns (previous run)
  - Add run ID in current IDs (prefilled with 0s)
  - Double check nork characters: å, ø, æ
  - Updat the 
  
## Tranfer folders

  - Move old folders to the respective section
  - Move foldres of seq ID to RECEN_RUN
  - 7z encryption in windows VM
  


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>



<!-- # ---------- # -->
<!-- # Extra code # -->
<!-- # ---------- # -->

```{r toggle, echo=FALSE, results='asis'}

cat('
<style>
.btn-toggle {
  background-color: #0275d8;
  border: none;
  color: white;
  padding: 8px 16px;
  font-size: 14px;
  border-radius: 12px;       /* Rounded corners */
  cursor: pointer;
  transition: background-color 0.3s ease;
  box-shadow: 0 2px 4px rgba(0,0,0,0.15);
}
.btn-toggle:hover {
  background-color: #025aa5;
}
</style>

<script>
function toggleFigure(id) {
  var fig = document.getElementById(id);
  if (fig.style.display === "none" || fig.style.display === "") {
    fig.style.display = "block";
  } else {
    fig.style.display = "none";
  }
}
</script>
')
```



```{r fill_template, echo = FALSE, eval = TRUE}

# Generate CheckList entries
html.output <- generate_checklist_html(checklist  = rev(cl.sections),
                                       categories = rev(cl.categories))

# Placeholder list
checklist.terms[['--runID--']] <- date.checklist
checklist.terms[['--checklist_buttons--']] <- html.output


# Fill CheckList template
template.to.fill   <- readLines(checklist.ready.file)

for (t in names(checklist.terms)) {
  
  template.to.fill <- gsub(x          = template.to.fill,
                           pattern     = t,
                           replacement = checklist.terms[[t]],
                           ignore.case = T, perl = T)
} 

# Export ready-to-use CheckList
writeLines(text = template.to.fill, con = checklist.ready.file)
```

## Email Drafts

<br>

### After launching *LocalApp*

<br>

```
Title: Bioinformatic analysis week WW

Dear all,

Sequencing data and all the metadata are now in TSD.
LocalApp analysis will start in an hour.

We will continue with TSOPPI and ICNVA on monday morning, so data for QC should be available shortly before Monday meeting.

Have a nice weekend

Jaime
```

<br>

### When *TSOPPI* and *ICNVA* are ready

<br>

```
Dear all,

TSOPPI and ICNVA analysis are running now.
The small variant tables should be ready for QC in ̃~30 minutes.

Best,

Jaime
```

<br>

### When reports are ready in TSD

<br>

```
Dear all, 

The reports and all the related data are now available in the //hypatia/InPred folder. We have sequenced and processed N IPA cases (N1_ID, N2_ID, ... ,NN_ID ). 

Sequencing went fine. No ctDNA data. 

We observed some sequencing-related warnings for IPA065, be aware that this may be reflected in the interpretation.

Best,

Jaime
```

<br>

## Q & A

<br>

### 1. How can I know the run ID `$RUN_ID` ?

  - The run ID is in the name of the sequencing files and it is automatically generated from the sequencer.
  
  - The run ID has the following format `YYMMDD_AAAAAAAA_BBBB_CCCCCCCCCC` (Figure 12). You can copy the prefix name (i.e., without the file extension) and save the content in the `$RUN_ID` variable.
  
  - In Figure 12, `RUN_ID="250606_NB552811_0087_AH2T7NBGYX"`.

<br>

```{r find_runid, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "find_runid",
  button.label = "Figure 12",
  img.src = "fig/12_Find_run_ID.png",
  fig.width = "95%",
  fig.cap = "The run ID is the prefix of the sequencing files, in this example, the file name without the *.md5* extension"
)
```

<br>

### 2. Visualize fusions with `FusViz`. 

<br>

  - Open a new terminal, but do not acces to the `Colossus` server.
  
  - Go to the *FusViz* folder (`$DURABLE/software/fusviz`) and launch the image.
  
  - Open the http link in a browser, in this example `http://127.0.0.1:5776`, (Figure 13).

```{bash, echo=TRUE, eval = FALSE}
# Launch image from fuzviz folder
cd $DURABLE/software/fusviz

singularity run fusviz_v1.0.0.sif 4000
```

```{r launch_fusviz, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "launch_fusviz",
  button.label = "Figure 13",
  img.src = "fig/13_Launch_fusviz_image.png",
  fig.width = "95%",
  fig.cap = "When launching the fusviz image, you will see the same text in the terminal"
)
```

<br>

  - Click On the *Home* tab (left-side). Select the genome *hg19 (offline)* and upload your text files with the SV details (Figure 14).
  
```{r load_SV, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "load_SV",
  button.label = "Figure 14",
  img.src = "fig/14_Load_SV_table.png",
  fig.width = "100%",
  fig.cap = "Select genome and load SV table"
)
```

<br>

  - Click On the *Linear* tab (left-side). Click on the *Upload a file* button, and add the `hg19.fasta.fai`, `hg19.fasta`, and `cytoBand.txt`, all of them available in the *fusviz* folder `$DURABLE/software/fusviz` (Figure 15).
  
  - Then click on the *Load gene track (offline)* button.

```{r load_genetrack, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "load_genetrack",
  button.label = "Figure 15",
  img.src = "fig/15_Load_genome.png",
  fig.width = "100%",
  fig.cap = "Load hg19 fasta, index and cytoBand files"
)
```

<br>
  
  - Click On the *Two-way* tab (left-side). Now you will see three tabs in the top. Click on the *Overview_plot* button. 
  
    - Manually select the two genes involved in the fusion and their breakpoints (position in the genome where the fusion occurred). These numbers are the same values provided in the SV table (Figure 16).
    
    - You will see a plot in the right side with all the transcripts associated to each of the selected genes (upstream: green; downstream: orange), the location in their chromosomes and the parts of the transcript connected through the fusion.
    
    - Click on the *Download plot* at the bottom to save the figure.

<br>

```{r overview_plot, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "overview_plot",
  button.label = "Figure 16",
  img.src = "fig/16_Overview_plot.png",
  fig.width = "100%",
  fig.cap = "The *Overview_plot* option is located in the *Two-way* tab in the left menu. Select the genes and their breakpoints manually (left side) to see the connected parts in the transcripts (right side). Upstream gene: green; Downstream gene: orange"
)
```

<br>

  - Click On the *Two-way* tab (left-side). Now you will see three tabs in the top. Click on the *Domain_plot* button. 
  
    - Manually select the two genes involved in the fusion, select one transcript for each gene and their breakpoints (position in the genome where the fusion occurred).
    
    - You will see a plot in the right side with the exons and annotated protein domains across the selected transcripts (upstream: green; downstream: orange). The arrow connecting the two transcripts indicates the place where the fusion occurs (Figure 17).
    
    - Click on the *Download plot* at the bottom to save the figure.

<br>

```{r domain_plot, echo=FALSE, results='asis'}
insert.fig.toggle(
  fig.id = "domain_plot",
  button.label = "Figure 17",
  img.src = "fig/17_Domain_plot.png",
  fig.width = "100%",
  fig.cap = "The *Domain_plot* option is located in the *Two-way* tab in the left menu. Select the genes, one transcript per gene and their breakpoints manually (left side) to see the exons and protein domains of the fused transcripts (right side). Upstream gene: green; Downstream gene: orange"
)
```

<br>

## To be added:

  - gpg --import /tsd/p1150/data/durable/gpg-keys/p1150-key1@tsd.usit.no-private-key.asc
  - Check which samplesheet has the invisible character
  - Add figure about How to get the sequence ID ?
  - SampleSheet_YYMMDD_info.csv ? Is this automatically generated or imported?
  - What does adapt TSOPII table if needed?
