---
title: "InPreD_README"
author: "Jaime Abraham Castro-Mondragon"
date: 'Last update: `r Sys.Date()`'
output: 
  html_document:
    number_sections: true
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    self_contained: true
    code_folding: show
---

```{r setup, include=FALSE}
library(emo)

knitr::opts_chunk$set(echo = TRUE)

date.checklist <- substr(gsub(x = Sys.Date(), pattern = "-", ""), start = 3, 8)

checklist.terms <- list()
```


```{r functions, include=FALSE}
# ----------------------- #
# Show/Hide figure botton #
# ----------------------- #

insert.fig.toggle <- function(
  fig.id,        # Unique id for the div/button (e.g. "figVMWare")
  button.label,  # Label for the button (e.g. "Toggle Figure 1")
  img.src,       # Path to image (e.g. "fig/01_VMWare_login.png")
  fig.width = "45%",  # Width style for image
  fig.cap = "Caption text"  # Figure caption text
) {
  safe_fig_id <- gsub("'", "\\'", fig.id)
  safe_button_label <- gsub("'", "\\'", button.label)
  safe_img_src <- gsub("'", "\\'", img.src)
  safe_fig_cap <- gsub("'", "\\'", fig.cap)
  
  cat(sprintf('
<div style="margin-bottom: 10px; text-align: right;">
  <button class="btn-toggle" onclick="toggleFigure(\'%s\')">%s</button>
</div>
<div id="%s" style="display:block; text-align: center; margin-bottom: 20px;">
  <img src="%s" style="width:%s;" class="img-fluid"/>
  <p><strong>%s.</strong> %s</p>
</div>
', safe_fig_id, safe_button_label, safe_fig_id, safe_img_src, fig.width, safe_button_label, safe_fig_cap))
}


# ------------------------------ #
# Checklist names and categories #
# ------------------------------ #
make_prepender <- function(prefix_name) {
  function(label, v) {
    c(label, v)
  }
}


add.cl.section    <- make_prepender("section")
add.cl.categories <- make_prepender("category")
```



```{r checklist, include=FALSE}

checklist.template.file <- "templates/InPreD_pipeline_Checklist.html"
checklist.ready.file    <- paste0(date.checklist, "_InPreD_pipeline_Checklist.html")

# ------------------------------- #
# Prepare CheckList section names #
# ------------------------------- #

# CheckList section names
cl.sections   <- vector()
cl.categories <- vector()

# Check List item
generate_checklist_html <- function(checklist, categories) {
  if (length(checklist) != length(categories)) {
    stop("Checklist and categories must have the same length.")
  }
  
  html_items <- mapply(function(item, i, category) {
    sprintf(
      '<div class="checklist-item %s">\n  <input type="checkbox" id="item%d">\n  <label for="item%d">%d. %s</label>\n</div>',
      category, i, i, i, item
    )
  }, checklist, seq_along(checklist), categories, USE.NAMES = FALSE)
  
  paste(html_items, collapse = "\n")
}


## Copy template and read copy
file.copy(from = checklist.template.file,
          to   = checklist.ready.file)

```



## Intro

This document explains how to run the InPreD pipeline to detect CNAs, driver mutations, gene fusions and other mutations from a Illumina TS0500 panel. Every week we sequence a batch of max 8 samples (hereafter referred as *sequencing run*). 

In this document we will use the following sequence run ID, we can easily identify by the first 6 digits corresponding to the data when the sequences were generated.

```{bash, echo=TRUE, eval = FALSE}
250515_AAAAAAAA_BBBB_CCCCCCCCCC
```

<!-- # Steps -->

<br>

## Connect to the Virtual Machines

<br>

In your browser, open the [VMWare Horizon](https://view.tsd.usit.no/portal/webclient/#/home) website. Log in with your *username* and *2FA* first, then VMWare will ask your *password* (Figure 1).

```{r figVMWare, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "figVMWare",
  button.label = "Figure 1",
  img.src = "fig/01_VMWare_login.png",
  fig.width = "45%",
  fig.cap = "Log in in the VMWare Horizon portal"
)

```

2. Click on the **p1150-rhel8-01-pool** icon, this will open the Linux VM (Figure 2). The other icon (*p1150-win02-pool*) will open the Windows VM.

```{r figVMs, echo=FALSE, results='asis'}

insert.fig.toggle(
  fig.id = "figVMs",
  button.label = "Figure 2",
  img.src = "fig/02_VMs.png",
  fig.width = "45%",
  fig.cap = "Available VMs. Left: Linux, Right: Windows"
)

```

<br>

## Folder structure

We will be working in the following directory: `/ess/p1150/data/durable`. Next we have a list of folders that will be constantly used during the analysis, from now on, **the file paths are relative to the working directory**.


```{bash, echo=TRUE, eval = FALSE}
cd /ess/p1150/data/durable
```

 <br>
 
  - **`file-import/p1150-member-group/tsd`** : this is the folder where raw files are available. The files are usually uploaded every friday.

<br>

  - **`file-export/`** : this is the folder where files can be exported outside the [TSD](https://selfservice.tsd.usit.no/) framework.

<br>

  - **`TSO500_SampleSheets/`** : in this folder we should place the Sample sheet files (see Step X) generated in every sequence run. Every sequence run should have three associated files. 
    - `250515_info.csv` : Original Samplesheet (for this sequencing run ID). 
    - `250515_info_TSOPPI.csv` : Customized Samplesheet (required by TSOPPI, may contain samples from many run IDs).
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC.csv` : Curated Samplesheet.

<br>

  - **`/TSO500_pronto_metadata/`** : in this folder we should place the metadata tables (original and in a compressed version; see Step X). Every sequence run should have two associated files. 
    - `PRONTO_metadata_250515.txt` : Original metadata. 
    - `PRONTO_metadata_250515.7z` : Compressed table.

<br>

  - **`TSO500_sequences/`** : in this folder we should place the sequence files (see Step X) generated in every sequence run. Every sequence run should have two associated files. 
    - `250515/` : Folder with sequences. 
    - `250515.md5` : File with checksums.

<br>

  - **`/TSO500_scripts_and_logs/`** : in this folder we will place the log and error messages of each analysis launched in the pipeline. One folder per sequence run ID.

<br>

  - **`TSO500_analyses/`** : here we will find the results after running *LocalApp*. Every sequence run should have one associated folder. 
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_LocalApp_analysis/` : this is a link. 
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_LocalApp_analysis_fromRAW/` : *LocalApp* results : 
        - `results/${SAMPLE_ID}/` : Sample ID results are in individual folders.
        - `results/MetricsOutput.tsv` : Table with sequencing metrics

<br>

  - **`TSO500_postprocessing/`** : here we will find the results after running *LocalApp* and *TSOPPI*, two folders per sequence run ID. Both folders follow the same structure.
  
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_LocalApp_postprocessing_results/` : *LocalApp*. 
        - `${SAMPLE_ID}` : One folder per sample.
        - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_metrics_plots` : Sequencing QC-related plots.
        - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_variant_summary.csv` : Table with variants.

    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC_TSO500_TSOPPI_v0.3.2_main_2022_06_07/`: *TSOPPI*.

<br>

  - **`TSO500_icnva/`** : here we will find the [*PureCN*](https://scfbm.biomedcentral.com/articles/10.1186/s13029-016-0060-z) results. One folder per sequence run ID, and one folder per sample ID.
  
    - `250515_AAAAAAAA_BBBB_CCCCCCCCCC/${SAMPLE_ID}_icnva/`
        - `PureCN_results_default/` : *PureCN* results of each patient. Including the local optima solutions for ploidy/purity.
        - `PureCN_results_custom/` : *PureCN* results of each patient.
        - `${SAMPLE_ID}_NRC_plots.pdf` : Tumor vs Normal samples comparison plots.

<br>

  - **`TSO500_export_for_interpretation/`** : this is the folder where the results are stored before the interpretation of variants, mutations and fusions.
  
    - `RECENT_RUN/` : here we will find the 8 samples corresponding to the current sequencing run. The content is similar to `TSO500_icnva/`.
    - `IPA001-IPA100` : there are many folders with a similar pattern in their name. After interpretation the `RECENT_RUN/` content must be moved to their corresponding folder.

<br>

  - **`software/`** : here we will find the bioinformatic analysis pipeline scripts and the soruce code for all the software required. 
    - **`pipeline/v003_upgrading_pronto_to_1.2.0`**

<br>

4. Go to the pipeline folder, the analysis must be launched from this path.
```{bash, echo=TRUE, eval = FALSE}
cd software/pipeline/v003_upgrading_pronto_to_1.2.0
```

<br>

## Prepare pipeline

Before launching the pipeline we must verify if there is enough space in the disk and the size of the sequencing data.

<br>

### Verify disk space availability

  - Disk space : we need at least 200GB of free space. If this is not the case, contact people involved in the project to release some space.

```{bash, echo=TRUE, eval = FALSE}
# Check disk space
df -h /ess/p1150
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Verify disk space availability", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Verify size of sequencing run

  - Size of files from current sequence run: usually a sequencing run produces a folder with size varying from ~30-60 GB. If this number is outside this range, please contact the people involved in sequencing.

```{bash, echo=TRUE, eval = FALSE}
# Check size of sequencing results
du -csh /ess/p1150/data/durable/file-import/p1150-member-group/tsd
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Verify size of sequencing run", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Connect to hpc server.

**NOTE**: we must be connected to this server to launch the pipeline. Data manipulation can be done outside.

```{bash, echo=TRUE, eval = FALSE}
# Connect to hpc
ssh p1150-hpc-01
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Connect to hpc server", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Pipeline run directory.

**NOTE**: the pipeline must be launched from this folder.

```{bash, echo=TRUE, eval = FALSE}
cd  /ess/p1150/data/durable/software/pipeline/v003_upgrading_pronto_to_1.2.0
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Pipeline run directory", cl.sections)
cl.categories <- add.cl.categories("prepare_pipeline", cl.categories)
```

<br>

### Set up global variables.

The following variables must be set up before launching the pipeline:

  - Sequencing run ID = `RUN_ID`
  - Input data type = `INPUT_DATA_TYPE`
  - Sequencing run date = `RUN_DATE`
  - Durable folder = `DURABLE_DIR`
  
```{bash, echo=TRUE, eval = FALSE}
RUN_ID="250515_AAAAAAAA_BBBB_CCCCCCCCCC"
INPUT_DATA_TYPE="RAW"
RUN_DATE="250515"
DURABLE_DIR="/tsd/p1150/data/durable"
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Set up  ${RUN_ID}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Set up  ${INPUT_DATA_TYPE}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Set up  ${RUN_DATA}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Set up  ${DURABLE_DIR}", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)
```

<br>

### Prepare sequences and metadata

<br>

The sequences and the *PRONTO* metadata for the new run are located in the `file-import` folder.
They must be moved to the `TSO500_sequences` and `TSO500_pronto_metadata` folder, respectively, before launching the analysis.

An invisible character is inserted in the header of the *PRONTO* metadata file. This character must be removed because the table cannot be properly read by other software and crash the analysis. Once this issue is solved, proceed to move the folders.


```{bash, echo=TRUE, eval = FALSE}
# Move sequences file
mv ${DURABLE_DIR}/file-import/p1150-member-group/SampleSheet_${RUN_DATE} ${DURABLE_DIR}/TSO500_sequences

# Move PRONTO metadata file
mv ${DURABLE_DIR}/file-import/p1150-member-group/tsd/${RUN_DATE}* ${DURABLE_DIR}/TSO500_pronto_metadata/PRONTO_metadata_${RUN_DATE}.7z
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Remove invisible character in PRONTO metadata", cl.sections)
cl.categories <- add.cl.categories("important", cl.categories)

cl.sections   <- add.cl.section("Move sequences files", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)

cl.sections   <- add.cl.section("Move PRONTO metadata", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

### Verify flags and source `config.sh`

Verify the three following flags in the config file.
  
```{bash, echo=TRUE, eval = FALSE}
# Default ';' but other separator (',', '-', '/') can be used
SEPARATOR=";"

DECRYPT_UNPACK_CHECKSUM="T"
REMOVE_DECRYPTED_COMPRESSED_DATA="T"
```

<br>

Then source the `config` file

```{bash, echo=TRUE, eval = FALSE}
source config.sh ${RUN_ID}
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("config.sh : SEPARATOR", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("config.sh : DECRYPT_UNPACK_CHECKSUM", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("config.sh : REMOVE_DECRYPTED_COMPRESSED_DATA", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Source config.sh", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

```

<br>

### Prepare folder structure

Prepare the folder structure for a new run, this will create all folders to store
results and logs.


```{bash, echo=TRUE, eval = FALSE}
bash 00_setup_folder_structure_for_new_run.sh ${RUN_ID}
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Prepare folder structure", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)
```

<br>

### Generate SampleSheet 

<br>

We will generate a sample sheet with metadata for the current run ID.

```{bash, echo=TRUE, eval = FALSE}
bash 00_generate_TSO500_sample_sheet.sh ${RUN_ID} ${INPUT_DATA_TYPE} \
  &> ${SCRIPTS_AND_LOGS_FOLDER}/00_generate_TSO500_sample_sheet.log
```

<br>

Copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 00_generate_TSO500_sample_sheet.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/00_config.sh
```

```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Generate sample sheet", cl.sections)
cl.categories <- add.cl.categories("setup", cl.categories)

cl.sections   <- add.cl.section("Copy Samplesheet (00) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

## Launch TSO500 pipeline

<br>

Now we are ready to launch the *TSO500* analysis pipeline. It has to be launched
in many steps, and many of them require manual intervention.

<br>

### Launch LocalApp analysis

<br>

This analysis performs the data QC for the sequencing results.

`r emo::ji("hourglass_flowing_sand")` Running time: ~13 hours

```{bash, echo=TRUE, eval = FALSE}
bash 01_raw_sequencing_data_management_and_TSO500_analysis.sh ${RUN_ID} ${INPUT_DATA_TYPE} \
  > ${SCRIPTS_AND_LOGS_FOLDER}/01_raw_sequencing_data_management_and_TSO500_analysis.stdout \
  2> ${SCRIPTS_AND_LOGS_FOLDER}/01_raw_sequencing_data_management_and_TSO500_analysis.stdout &
```

<br>

Send an email to the local *InPreD* team to to tell them that the analyses have
been launched.

<br>

Once this analysis is done, copy the logs and scripts for this sequencing run in
the `${SCRIPTS_AND_LOGS_FOLDER}` folder.

```{bash, echo=TRUE, eval = FALSE}
cp -a 01_raw_sequencing_data_management_and_TSO500_analysis.sh ${SCRIPTS_AND_LOGS_FOLDER}
cp -a config.sh ${SCRIPTS_AND_LOGS_FOLDER}/01_config.sh
```


```{r, echo=FALSE, eval = TRUE}
cl.sections   <- add.cl.section("Launch LocalApp", cl.sections)
cl.categories <- add.cl.categories("analysis", cl.categories)

cl.sections   <- add.cl.section("Send LocalApp email", cl.sections)
cl.categories <- add.cl.categories("email", cl.categories)

cl.sections   <- add.cl.section("Copy LocalApp (01) scripts and logs", cl.sections)
cl.categories <- add.cl.categories("file_manip", cl.categories)
```

<br>

### Launch TSOPPI analysis

<br>

This analysis performs different analysis included in [TSOPPI](https://tsoppi.readthedocs.io/en/latest/index.html), such as:

 - Quality control (QC) metrics for samples
 - Fusion and splice variant output for RNA samples;
 - Small variant output, copy number variant output (CNV) and microsatellite status (MS status) information for DNA samples.

<br>

TSOPPI:

 - Update table manually, insert figure here
 - Update ALL_RUN_IDs variable in script (see README)
 - Launch it

<br>
<br>
<br>


<!-- # ---------- # -->
<!-- # Extra code # -->
<!-- # ---------- # -->

```{r toggle, echo=FALSE, results='asis'}

cat('
<style>
.btn-toggle {
  background-color: #0275d8;
  border: none;
  color: white;
  padding: 8px 16px;
  font-size: 14px;
  border-radius: 12px;       /* Rounded corners */
  cursor: pointer;
  transition: background-color 0.3s ease;
  box-shadow: 0 2px 4px rgba(0,0,0,0.15);
}
.btn-toggle:hover {
  background-color: #025aa5;
}
</style>

<script>
function toggleFigure(id) {
  var fig = document.getElementById(id);
  if (fig.style.display === "none" || fig.style.display === "") {
    fig.style.display = "block";
  } else {
    fig.style.display = "none";
  }
}
</script>
')
```



```{r fill_template, echo = FALSE, eval = TRUE}

# Generate CheckList entries
html.output <- generate_checklist_html(checklist  = rev(cl.sections),
                                       categories = rev(cl.categories))

# Placeholder list
checklist.terms[['--runID--']] <- date.checklist
checklist.terms[['--checklist_buttons--']] <- html.output


# Fill CheckList template
template.to.fill   <- readLines(checklist.ready.file)

for (t in names(checklist.terms)) {
  
  template.to.fill <- gsub(x          = template.to.fill,
                           pattern     = t,
                           replacement = checklist.terms[[t]],
                           ignore.case = T, perl = T)
} 

# Export ready-to-use CheckList
writeLines(text = template.to.fill, con = checklist.ready.file)
```


## To be added:

  - gpg --import /tsd/p1150/data/durable/gpg-keys/p1150-key1@tsd.usit.no-private-key.asc
  - Check which samplesheet has the invisible character
  - Add figure about How to get the sequence ID ?
  - SampleSheet_YYMMDD_info.csv ? Is this automatically generated or imported?
  - 


<!-- # ----------- # -->
<!-- # Draft Email # -->
<!-- # ----------- # -->

Sequencing data and all the metadata are now in TSD and LocalApp analysis will start in an hour or so.

We will continue with TSOPPI and icnva on monday morning together with Jaime, so data for qc should be available shortly before Monday meeting or then around.
