---
title: "Notes about QC for small variants"
author: "Jaime Abraham Castro-Mondragon"
date: 'Last update: `r Sys.Date()`'
params:
    args: ''
output:
  html_document:
    number_sections: true
    theme: cosmo
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
    self_contained: true
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: '4'
---

```{r setup, include=FALSE}
library(emo)
library(glue)
library(broom)
library(klippy)

knitr::opts_chunk$set(echo = TRUE)

```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(lang     = c('r', 'python', 'bash'),
               position = c('top', 'right'),
               color    = '#253494')
```

<br>

## Introduction

<br>

After we run *LocalApp* and *TSOPPI* we have a list of small variants predicted
for each sample in a given `$RUN_ID`. These results have to be manually inspected
and curated to discard potential artifacts, variants in problematic regions (commonly
referred to as *black listed*) and germ line variants.

<br>

Steps:

- General QC
- Check *exclude* category variants
  - VAF
  - TSO500 category
  - Frequencies
- Check *include* category variants
  - VAF
  - TSO500 category
  - Frequencies
- Concordance DNA-RNA
- Gender
- Tumor content

## Prepare xlsx file

- First we create a copy of the *.tsv* file in excel, save this copy as *.xlsx* format,
all the updated should be done in this new file, and we never modify the *.tsv* file.

- Delimited, tab, text

- Excel Workbook
  - `${sample_id}_small_variant_table.tsv` to `${sample_id}_small_variant_table_forQC.xlsx`


- Open *IGV* and load the *.bam* files


## QC

### Evaluate the *exclude* category variants

We check the variants th

  - Variants that are not in the coding regions are automatically excluded
  - Check *TSO500* column with variant categories
    - Check also *Class_judgement*, *Depth*
  - Variants in the category *BL* are automatically excluded
  - Variants in the category *GL_DB* are automatically excluded: these variables
  are annotated in verified databases, we don't need to verify it.


### Evaluate the *include* category variants 

- Examples:
  - SNV (Single Nucleotide Variants)
  - Deletions (1-15)
  - Insertions (1-)
  
- We check the frequencies are very low in normal and hight in the sample.

- Variants that are not in the coding regions and synonymous are automatically excluded

- IGV:

  - Be sure to activate the option: *View as pairs*
  - On right click: *Sort alignment by* -> *Base*
  - This allows to see all the nucleotides matching the coordinates of the variant
  - The base we see are those mismatching the reference


## VAF 

- If the tumor content is high and the VAF of a variant is high, we cannot be sure
if this is a germn line variant

- Separation in VAF plot
  - We expect a big separation when the tumor content is high
  - Little separation is associated to low tumor content, because the sample comes
    from normal cells and we should expect normal VAF of ~0.5


## Concordance

- File: `${sample_id}_sample_concordance.txt`

Always verify the DNA-RNA reads concordance, if it is less than 80%, report it.

However we always check the RNA quality data (metrics), it is likely that this low
concordance value is due to one sample RNa or DNA of good quality and the other
one with low quality.

If we have both DNA and RNA with good quality and the concordance is low, then we
have a real problem.


## Gender prediction

See copy number plot and *PRONTO metadata*, they have to agree.

Otherwise, repor it.


## Tumor content

Plot vs PureCN

The predicted number by the patologist should not vary much from the top *PureCN* solutions


## Metrics summary plots

- Check contaminated samples
  - When the tumor content is high and many GL variants have low VAF (<=0.2) the
  sample may be tagged as contaminated becuase the software may assume there are
  is material from two different samples. Therefore this is not a true contamination
  
- Prioritize in coverage over actual numbers of reads
  



<!-- ```{r runID, echo=FALSE, eval = TRUE, comment=""} -->
<!-- cat(run.id) -->
<!-- ``` -->
